Reproducible sausage network buffers for built environment measurement: an open source GIS approach

### 2015-10-03 article/article.tex
PMH 20151002, again
PMH 20151002, again

% Relevance to planning and epidemiology
Developing geoprocessing methods around free and open source tools
holds particular benefits for disciplines such as epidemiology, public
policy, and urban planning because of the reduced barriers of adoption
for practitioners and advocates outside of larger institutions. Academic
institutions and larger public agencies may be able to afford costly software
applications. However, smaller agencies or advocacy groups wishing to
perform quantitative analysis of built environment might be unable to
make use of even transparently described analyses that are reliant on proprietary software, simply due to the
cost of software licensing fees.

% Not a literature review, per se. Since we are developing a methods
% paper largely in response to one particularly important method, I'm
% not sure how important a full-blown review is. Open to suggestions,
% however. This could potentially be split off and developed into
% something more complete. 
% PMH: I agree, we aren't doing an evaluation of the use of PostGIS in research.
Finally, the PostGIS platform is especially beneficial in that it
naturally facilitates reproducible, scripted analyses, as well as 
provides the core infrastructure for data storage and management on even very large research
projects. \textcite{hurvitz2014emerging} describe the use of
PostgreSQL/PostGIS to manage and process geospatial and timeseries
data from a variety of sources including Global Positioning Systems (GPS) and
accelerometers for very large data sets. However, we are neither first nor alone in identifying
PostGIS for use in research. A keyword search for ``PostGIS'' using
ISI Web of Science identified 35 \textcolor{red}{[I got 50 today!]} articles across a number of
disciplines describing the use of PostGIS in research. In the
transportation sector, \textcite{Wang2015routable} also used PostGIS
to store and analyze a large number of GPS traces in order to infer
characteristics about roadways such as position and rules, for the
purposes of generating routable roadway networks. In another recent
article, \textcite{Brovelli2015FOSS} describe the use of PostGIS
together with GeoServer to collect data about roadway pavement
conditions from the general public. Notable in this study is the
potential for PostGIS in supporting digitally-mediated public
participation. However, a PubMed search for ``PostGIS'' resulted in only one paper \parencite{Rowlingson2013}. In contrast, Web of Science and PubMed searches for ``ArcGIS'' resulted in 1,674 and 394 citations, respectively. This indicates that a number of researchers who could benefit from the use of PostGIS may simply be unaware of its existence.

\section*{Methods}
% Set-up and software pgRouting 2.x on OpenBSD 5.7 i5
Following the process outlined in 
\textcite{Forsyth2014sausage} and described in step-by-step detail in \textcite{ Forsyth2012proto}, we implemented the
sausage buffering technique using PostgreSQL 
with PostGIS and the pgRouting \parencite{pgRoutingContributors2013} extension, which implements a number of network routing algorithms, including Dijkstra's shortest path \parencite{Dijkstra1959}, which is also using in the ArcGIS Network Analyst exension. Using that technique we
generated 200 network buffers around randomly selected points. We then
compared those results to buffers analogously created in ArcGIS Desktop
10.2.X, for use as a basis for comparison with the results presented
by \citeauthor{Forsyth2014sausage}. Table \ref{tab:software} lists
software versions used in this analysis.

@@

%\subsection*{Data sources and network import}
To create the routable network required for generating the buffers, we
imported the Washington State portion of the Open Street Map (OSM) (\copyright\ OpenStreetMap contributors)
into PostGIS / pgRouting using the open source osm2pgrouting
utility. \textcolor{red}{Did we use GeoFabrik? If so we should describe.} Because much of our work deals with pedestrian travel, we
excluded highway facilities, which often do not allow for pedestrian
access. The osm2pgrouting import strategy yielded two primary
benefits. First, osm2pgrouting facilitates reproducible data
importation through the use of an XML-based configuration file, which 
allows facility exclusions to be carried out consistently, such as our exclusion of highways as non-traversible for
pedestrians. Secondly, osm2pgrouting
preserves the topology of the OSM network, rather than relying on
topology from the data set's simple vector geometry. Maintaining network topology is important for enforcing physical constraints such as over/underpasses or bridges.

@@

The PL/PgSQL implementation was used to generate buffers originating from 200 randomly
selected points, which were selected from the road data set's network vertices to ensure that they were not located in
roadless areas of the state. Then, in order to compare buffers
created using the PostGIS implementation with ArcGIS-generated buffers, we exported the origin points and OSM network data as shapefiles and followed the protocols described in
\textcite{Forsyth2012proto}, but using rounded buffer ends as described in \textcite{Forsyth2014sausage}.

@@

\section*{Results}
Overall, the differences between the PostGIS and ArcGIS generated buffers were quite low. Figure
\ref{fig:area_dens} shows the distribution of areas of these two sets of
buffers. Both exhibit the same somewhat bimodal shape, with greater
differences observed among smaller buffers. Although the PostGIS-generated buffers were slightly smaller, the mean difference's 95\% confidence interval was -0.23 to 0.04, indicating no significant difference in size using the two methods.

@@

The symmetric differences between the two sets of
buffers are small (figure \ref{fig:symdiff_dens}), \textcolor{red}{with X being less than Y}. \textcolor{cyan}{Can we add some numbers about the relative size of the differences to the size of the buffers?}. Figure \ref{fig:symmetric_differences} illustrates
three scenarios including the best, typical (closest to the median),
and worst cases with respect to these symmetric differences. In the
best case, the buffers encompassing a very small, isolated
pedestrian facility are practically identical. In the more typical
case, slight differences are visible upon close inspection. Finally,
in the worst case, which was located in a remote portion of Central
Washingtion, the buffers are very different, with
the PostGIS buffer only a subset of the larger ArcGIS-generated buffer.

@@

\section*{Discussion}
This study presented a new method for generating sausage buffers using free and open source GIS software, and compared the results of the new method to those of an existing method that relies on the leading proprietary GIS application, ArcGIS.

There were measurable, but not signiticant differences between buffers generated using the two methods. The primary source of differences between the PostGIS
and ArcGIS generated buffers is likely the result of topological differences
and errors in the network dataset. Osm2pgrouting imports OSM data into PostGIS while preserving network topology. Calculations performed with
PostGIS / pgRouting will properly handle non-apparent discontinuities,
such as the aforementioned cases of over/underpasses. The network used for creation of buffers in ArcGIS, on the other hand,
was exported as a shapefile from PostGIS. Because shapefiles do not preserve topology,
ArcGIS is forced to infer topology from simple ``spaghetti'' geometry. Thus, in the 
situation where a bridge crosses over another roadway, the
reconstructed network may erroneously connect between the bridge and
the road below. \textcolor{red}{Can we verify this?} In that case, the PostGIS
buffer would be smaller than the ArcGIS buffer, due to the greater, but erroneous 
connectivity in the ArcGIS network. While the mean difference in area
between our two sets of buffers was not statisitcally significant, the observed differences in our sample are consistent with the
logic that the more connected, inferred network topology would yield
larger buffers. \textcolor{red}{OK, sorry to ask, but we really should export back from shapefile to PostGIS as spaghetti for at least one of the bad cases of difference to test whether that was indeed the case. Or examine the roads in that worst case to verify that was happening. Then we needn't rely on a logical argument but can point directly to the cause.}

@@

In addition to establishing a reproducible method for creating sausage
buffers, the buffers created using PostGIS and ArcGIS 
are generally comparable. Differences between sausage buffers created in the two
programs are more likely to be due to differences in how network
datasets are constructed, than to differences in the underpinnings of how the buffers are
computed.

>>>>

% Relevance to planning and epidemiology
Developing geoprocessing methods around free and open source tools
holds particular benefits for disciplines such as epidemiology, public
policy, and urban planning because of the reduced barriers to adoption
for practitioners and advocates outside of larger institutions. Academic
institutions and larger public agencies may be able to afford costly software
applications. However, smaller agencies or advocacy groups wishing to
perform quantitative analysis of built environment might be unable to
make use of even transparently described analyses that are reliant on proprietary software, simply due to the
cost of software licensing fees.

% Not a literature review, per se. Since we are developing a methods
% paper largely in response to one particularly important method, I'm
% not sure how important a full-blown review is. Open to suggestions,
% however. This could potentially be split off and developed into
% something more complete. 
% PMH: I agree, we aren't doing an evaluation of the use of PostGIS in research.
Finally, the PostGIS platform is especially beneficial in that it
naturally facilitates reproducible, scripted analyses, as well as 
provides the core infrastructure for data storage and management on even very large research
projects. \textcite{hurvitz2014emerging} describe the use of
PostgreSQL/PostGIS to manage and process geospatial and timeseries
data from a variety of sources including Global Positioning Systems (GPS) and
accelerometers for very large data sets. We are admittedly neither first nor alone in identifying
PostGIS for use in research. A keyword search for ``PostGIS'' using
ISI Web of Science identified 35 \textcolor{red}{[I got 50 today!]} articles across a number of
disciplines describing the use of PostGIS in research. In the
transportation sector, \textcite{Wang2015routable} also used PostGIS
to store and analyze a large number of GPS traces in order to infer
characteristics about roadways such as position and rules, for the
purposes of generating routable roadway networks. In another recent
article, \textcite{Brovelli2015FOSS} describe the use of PostGIS
together with GeoServer to collect data about roadway pavement
conditions from the general public. Notable in that study is the
potential for PostGIS in supporting digitally-mediated public
participation. However, a PubMed search for ``PostGIS'' resulted in only one paper \parencite{Rowlingson2013}. In contrast, Web of Science and PubMed searches for ``ArcGIS'' resulted in 1,674 and 394 citations, respectively. This indicates that a number of researchers who could potentially benefit from the use of PostGIS may simply be unaware of its existence.

\section*{Methods}
% Set-up and software pgRouting 2.x on OpenBSD 5.7 i5
Following the process outlined in 
\textcite{Forsyth2014sausage} and described in step-by-step detail in \textcite{ Forsyth2012proto}, we implemented the
sausage buffering technique using PostgreSQL 
with PostGIS and the pgRouting \parencite{pgRoutingContributors2013} extension, which implements a number of network routing algorithms, including Dijkstra's shortest path \parencite{Dijkstra1959}, which is also used in the ArcGIS Network Analyst exension. Using that technique we
generated 200 sausage buffers around randomly selected points. We then
compared those results to buffers analogously created in ArcGIS Desktop
10.2.X, for use as a basis for comparison with the results presented
by \citeauthor{Forsyth2014sausage}. Table \ref{tab:software} lists
software versions used in this analysis.

@@

%\subsection*{Data sources and network import}
To create the routable network required for generating the buffers, we
imported the Washington State portion of the Open Street Map (OSM) (\copyright\ OpenStreetMap contributors)
into PostGIS / pgRouting using the open source osm2pgrouting
utility. \textcolor{red}{Did we use GeoFabrik? If so we should describe.} Because much of the work in measuring built environment exposures deals with pedestrian travel, we
excluded highway facilities, which often do not allow for pedestrian
access. The osm2pgrouting import strategy yielded two primary
benefits. First, osm2pgrouting facilitates reproducible data
importation through the use of an XML-based configuration file, which 
allows facility exclusions to be carried out consistently, such as our exclusion of highways as non-traversible for
pedestrians. Secondly, osm2pgrouting
preserves the topology of the OSM network, rather than relying on
topology created from the data set's simple vector geometry. Maintaining network topology is important for enforcing physical constraints to traversability such as over/underpasses or bridges.

@@

The PL/PgSQL implementation was used to generate buffers originating from 200 points, which were randomly selected from the road data set's network vertices to ensure that they were not located in
roadless areas of the state. Then, in order to compare buffers
created using the PostGIS implementation with ArcGIS-generated buffers, we exported the origin points and OSM network data as shapefiles and followed the protocols described in
\textcite{Forsyth2012proto}, but using rounded buffer ends as described in \textcite{Forsyth2014sausage}.

@@

\section*{Results}
Overall, the differences between the PostGIS and ArcGIS generated buffers were quite low. Figure
\ref{fig:area_dens} shows the distribution of areas of these two sets of
buffers. Both exhibit the same bimodal shape, with greater
differences observed among smaller buffers. Although the PostGIS-generated buffers were slightly smaller, the mean difference's 95\% confidence interval was -0.23 to 0.04, indicating no significant difference in size using the two methods.

@@

The symmetric differences between the two sets of
buffers were small (figure \ref{fig:symdiff_dens}), \textcolor{red}{with X being less than Y}. \textcolor{cyan}{Can we add some numbers about the relative size of the differences to the size of the buffers?}. Figure \ref{fig:symmetric_differences} illustrates
three scenarios including the best, typical (closest to the median),
and worst cases with respect to these symmetric differences. In the
best case, the buffers encompassing a very small, isolated
pedestrian facility are practically identical. In the more typical
case, slight differences are visible upon close inspection. Finally,
in the worst case, which was located in a remote portion of Central
Washingtion, the buffers are very different, with
the PostGIS buffer being only a subset of the larger ArcGIS-generated buffer.

@@

\section*{Discussion}
This study presented a new method for generating sausage buffers using free and open source GIS software, and compared the results of this method to those of an existing method that relies on the leading proprietary GIS application, ArcGIS.

There were measurable, but not signiticant differences between buffers generated using the two methods. The primary source of differences between the PostGIS
and ArcGIS generated buffers is \textcolor{red}{likely} the result of topological differences
and errors in the network dataset. Osm2pgrouting imports OSM data into PostGIS while preserving network topology. Calculations performed with
PostGIS / pgRouting will properly handle not readily apparent discontinuities,
such as the aforementioned cases of over/underpasses. The network used for creation of buffers in ArcGIS, on the other hand,
was exported as a shapefile from PostGIS. Because shapefiles do not preserve topology,
ArcGIS is forced to infer topology from simple ``spaghetti'' geometry. Thus, in the 
situation where a bridge crosses over another roadway, the
reconstructed network may erroneously connect between the bridge and
the road below. \textcolor{red}{Can we verify this?} In that case, the PostGIS
buffer would be smaller than the ArcGIS buffer, due to the greater, but erroneous 
connectivity in the ArcGIS network. While the mean difference in area
between our two sets of buffers was not statisitcally significant, the observed differences in our sample are consistent with the
logic that the more connected, inferred network topology would yield
larger buffers. \textcolor{red}{OK, sorry to ask, but we really should export back from shapefile to PostGIS as spaghetti for at least one of the bad cases of difference to test whether that was indeed the case. Or examine the roads in that worst case to verify that was happening. Then we needn't rely on a logical argument but can point directly to the cause.}

@@

In addition to establishing a reproducible method for creating sausage
buffers, the buffers created using PostGIS and ArcGIS 
are generally comparable. Differences between sausage buffers created in the two
programs are more likely due to differences in how network
datasets are constructed, than to differences in the underpinnings of how the buffers are
computed.
